---
title: "Clinical Survival Analysis Using survkit"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clinical Survival Analysis Using survkit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8
)
```

## Overview

This vignette demonstrates how to perform time-to-event analysis using the survkit package. The goal is to generate Kaplan-Meier and Cox-adjusted survival curves with appropriate handling of censoring and uncertainty.

Survival analysis is commonly used in clinical research to evaluate outcomes such as time to disease progression, relapse, or death.

## Research Objective

To evaluate differences in survival outcomes between groups and assess how covariates influence time-to-event outcomes using both unadjusted and adjusted survival models.

## Data Description

We use the `lung` dataset from the survival package.

This dataset includes:

- `time`: survival time in days
- `status`: event indicator (1 = censored, 2 = death)
- `sex`: grouping variable (1 = male, 2 = female)
- `age`: patient age
- `ph.ecog`: ECOG performance status score

For analysis, the event indicator is recoded to a binary variable:

- 1 = event occurred (death)
- 0 = censored

## Load Libraries
```{r setup}
library(survkit)
library(survival)
```

## Data Preparation
```{r data-prep}
data(lung)

# Recode status: 2 (death) -> 1 (event), 1 (censored) -> 0
lung$status <- ifelse(lung$status == 2, 1, 0)

# Check data structure
head(lung[, c("time", "status", "sex", "age", "ph.ecog")])
```

## Kaplan-Meier Survival Analysis

Kaplan-Meier estimation is used to compute survival probabilities over time without adjusting for covariates.
```{r km-plot, fig.width=10, fig.height=8}
km_plot <- survkit(
  lung,
  "time",
  "status",
  "sex",
  title = "Kaplan-Meier Survival Curve",
  subtitle = "Unadjusted Survival by Sex"
)

km_plot
```

### Interpretation

The Kaplan-Meier curves represent the probability of survival over time for each group.

- Step decreases indicate observed events (deaths)
- Censor marks (+ symbols) indicate incomplete follow-up
- Differences between curves suggest variation in survival between groups
- Risk table shows number of patients remaining at risk at each timepoint

## Cox-Adjusted Survival Analysis

Cox proportional hazards models allow adjustment for additional covariates.
```{r cox-plot, fig.width=10, fig.height=8}
cox_plot <- survkit(
  lung,
  "time",
  "status",
  "sex",
  adjust_for = c("age", "ph.ecog"),
  title = "Cox-Adjusted Survival Curve",
  subtitle = "Adjusted for Age and ECOG Performance Status"
)

cox_plot
```

### Interpretation

The adjusted curves represent survival estimates at reference covariate values (mean age, mean ECOG score) derived from the Cox model.

- Curves are smoother because they represent model-based predictions
- Confidence intervals reflect uncertainty in estimated survival probabilities
- Censor marks are not shown, as they are not defined for predicted curves
- Risk table shows observed patient counts from the complete-case analysis

## Comparison: Unadjusted vs Adjusted
```{r comparison, fig.width=16, fig.height=8}
library(patchwork)

km_plot | cox_plot
```

### Interpretation

Comparing unadjusted and adjusted curves helps identify whether observed group differences are explained by other variables.

In this example:

- Both plots show females (sex = 2) have better survival than males (sex = 1)
- The difference persists after adjusting for age and performance status
- This suggests the survival difference is not fully explained by these covariates

## Customization

survkit allows extensive customization of visual elements:
```{r custom-plot, fig.width=10, fig.height=8}
custom_plot <- survkit(
  lung,
  "time",
  "status",
  "sex",
  line_width = 2,
  censor_size = 4,
  conf_int_alpha = 0.3,
  viridis_option = "plasma",
  title = "Customized Survival Curve"
)

custom_plot
```

## Statistical Considerations

- **Kaplan-Meier estimates** are non-parametric and based on observed data
- **Cox models** estimate hazard ratios while adjusting for covariates
- **Censoring** must be handled correctly to avoid biased estimates

survkit enforces these principles by:

- Showing censor marks only for observed data (not predictions)
- Representing uncertainty using confidence intervals
- Aligning risk tables with observed patient counts
- Making statistically appropriate decisions automatically

## Saving Publication-Ready Figures
```{r save-plot, eval=FALSE}
# Save at 600 DPI with golden ratio dimensions
save_survkit("km_survival.png", km_plot, width = 10, dpi = 600)
save_survkit("cox_survival.png", cox_plot, width = 10, dpi = 600)
```

## Conclusion

This example demonstrates how survkit can be used to produce statistically sound and interpretable survival visualizations for clinical research.

The ability to generate both unadjusted and adjusted survival curves with integrated risk tables supports reproducible workflows and improves interpretability of time-to-event analyses.

### Key Features

1. **Unified interface** for KM and Cox-adjusted curves
2. **Automatic risk tables** perfectly aligned with curves
3. **Statistically rigorous** handling of censoring and uncertainty
4. **Publication-ready styling** with customization options
5. **Reproducible** workflows for clinical research

## References

- Cox, D. R. (1972). Regression Models and Life-Tables. *Journal of the Royal Statistical Society: Series B*, 34(2), 187-202.
- Kaplan, E. L., & Meier, P. (1958). Nonparametric estimation from incomplete observations. *Journal of the American Statistical Association*, 53(282), 457-481.
- Therneau, T. M. (2023). *A Package for Survival Analysis in R*. R package version 3.5-7.
